Verification of Stencil Computations through Spatial Specifications

### 2016-07-07 paper.tex
edits to eval and intro

\begin{abstract}
  Verifying the correctness of numerical computations relies on first
  giving a specification of their behaviour. In many cases, a
  numerical computation is derived from an underlying mathematical
  model, but this is often difficult to use as a specification as it
  is highly decoupled from the implementation. We have been studying
  specification techniques in the middle ground between low-level and
  high-level specifications for numerical code. In this paper, we
  focus on \emph{stencil computations}---a common idiom in numerical
  code, but one which is error-prone due to fine-grained indexing
  errors. We hypothesise that in practise, stencil computations tend
  to have a regular shape which is amenable to static analysis and
  simple specification. We describe a abstract spatial specification
  language for stencil computations that can be embedded as
  annotations in source code, along with inference, checking, and
  specification and program synthesis procedures. We evaluate our
  language against a corpus of numerical Fortran code, and show that
  the vast majority of stencil computations indeed have a simple,
  regular, static shape. Around 10\% of the stencils found evoke a
  programming pattern for which indexing errors are plausible. We
  investigate two stencils in detail and find that we are able to
  detect the vast majority of potential errors. This paper details our
  stencil specification language, its static semantics, inference,
  checking, and program synthesis procedures, our implementation, and
  our evaluation study.
\end{abstract}

@@

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
(b) Inferred and synthesised specification for (a) by CamFort.
\vspace{0.5em}
\hrule
%
\begin{center}
\begin{tikzpicture}[scale=0.85]
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
\vspace{-0.5em}
(c) Pictorial representation of the two stencil specifications.
The horizontal dimension is \texttt{dim=1} and the vertical is \texttt{dim=2}.
\caption{Fragment of a Navier-Stokes fluid simulator and its
  specification in our language}
\label{ref:navier-stokes-fragment}
\vspace{-0.8em}
\end{figure}

@@

\renewcommand*{\arraystretch}{0.8}
\paragraph{Target language syntax} Throughout, for the source language
syntax, $e$ ranges over its expressions (which may be impure).  By ``variable'', we mean
in the imperative sense (\ie{}, named binders to memory cells) instead
of mathematical variables.

\begin{definition}[Base induction variable]
  A variable \textit{v} of integer type is a \emph{base induction
    variable} if it is the control variable of a ``for''
  loop,\footnote{Or equivalent, \eg{}, \texttt{do} in Fortran, for our implementation.} incremented by $1$ per
iteration. The variable is marked as a base induction variable
only within the scope of the loop. Induction variables are ranged
over by $i$, $j$, $k$ throughout. Since we do not use non-base
(derived) induction variables we often simply say \emph{induction variable}.
\end{definition}

@@

\begin{figure}[t]
\input{region-model.tex}
%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
%  the multi-set $A$}
\caption{Model of regions,
$\interp{-}_n : \textit{region} \rightarrow \mathcal{P}(\domainVal{}^n)$}
\label{fig:region-model}
\vspace{-0.7em}
\end{figure}

@@

\begin{definition} \textsf{Mult}
and \textsf{Approx} are parametric labelled variant types
with injection function given by their definition:
%
\begin{align*}
\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
\end{align*}
\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
on. The $\textsf{Mult}$ type corresponds to the presence or absence of the
\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}. 
The $\textsf{Approx}$ type corresponds to the presence or absence of
the spatial approximation modifier, with \textsf{exact} when there is
no such modifier and \textsf{lower} and \textsf{upper} for
\term{atLeast} and \term{atMost}. Later, it will be useful to model
pairs of lower and upper bounds which is provided by \textsf{both} 
(but which isn't used here). These are used for labelled models, and other components later in
Section~\ref{sec:analysis}.% with additional information.
\label{def:mult-and-approx}
\end{definition}

\begin{figure}
\input{spatial-model.tex}
\caption{\textit{spec} model,
$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
\label{fig:spatial-model}
\vspace{-0.75em}
\end{figure}

\subsection{Soundness}

Given the above model, we can
consider soundness for the equational theory of
specifications and the notion of spatial approximation. The proofs are given in our supplement.

@@

\begin{definition}[Approximation]
A specification $S'$
\emph{approximates} a more specific specification $S$, written 
$S \sqsubseteq S'$, \emph{iff}:
$$
\forall M_c \in \textsf{Mult}(\domainVal{}^n) . \mathit{cons}_n(\interp{S}_n,M_c) \implies
\mathit{cons}(\interp{S'}_n,M_c)
$$
\end{definition}
%
For the syntactic sub-specification equations on regions $<:^r$, it follows that
these imply semantic approximation:
%
\begin{lemma}[Sub-specification $<:^r$ coincides with approx 
$\sqsubseteq$]
\begin{equation*}
\forall S, S' . \; S <:^r S' \Rightarrow S \sqsubseteq S'
\end{equation*}
\end{lemma}
%
However, the converse direction does not hold. 
The regions describing the more acceptiong five-point stencil
above do not follow from the definition of $<:$ on the previous
more specific specification.%, that is:
%${\textcolor{darkgreen}{\term{centered(depth=1, dim=1)*reflexive(dim=2)
%      + centered(depth=1, dim=2)*reflexive(dim=1)}}}
%\not{<:} {\textcolor{darkgreen}{\term{centered(depth=1,dim=1) +
%      centered(depth=1,dim=2)}}}$.
 However, consitency (and thus $\sqsubseteq$) shows
that we can check the five-point stencil code against the more general
specifcation of {\textcolor{darkgreen}{\term{fivepoint}}}. Thus, we
can always use an approximation in place of the
more specific specification. Although this decreases the likelihood of catching
stencil errors (as more access patterns conform), it reduces the space and
effort it takes to write the specification.

\subsection{Synthesis\label{subsec:synthesis}}

\section{Evaluation}
\label{sec:evaluation}
We evaluated the stencil language on a corpus of scientific code. We first give an overview of the corpus itself before presenting our results.

@@

Package   & {Total} & {Parsed} & {Total} & {Parsed}  \\ % braces prevent headers from being formatted as numbers
\hline
UM        & \umFiles                       & \umparseOk                      & \umLoC                      & \umlinesParsed \\
E3MG      & \ethreemgeaFiles               & \ethreemgeaparseOk              & \ethreemgeaLoC              & \ethreemgealinesParsed \\
BLAS      & \blasFiles                     & \blasparseOk                    & \blasLoC                    & \blaslinesParsed \\
Hybrid4   & \hybridfourFiles               & \hybridfourparseOk              & \hybridfourLoC              & \hybridfourlinesParsed \\
GEOS-Chem & \geoschemFiles                 & \geoschemparseOk                & \geoschemLoC                & \geoschemlinesParsed \\
Navier    & \navierFiles                   & \navierparseOk                  & \navierLoC                  & \navierlinesParsed \\
CP        & \computationalphysicstwoFiles  & \computationalphysicstwoparseOk & \computationalphysicstwoLoC & \computationalphysicstwolinesParsed \\
\hline
\textbf{Total} & \overallFiles  & \overallparseOk & \overallLoC & \overalllinesParsed \\
\hline
\end{tabular}
\caption{Summary of software packages used for evaluation\label{tab:corpus}}
\end{table}

Table~\ref{tab:corpus} shows summary statistics of the software
packages used in our evaluation, all of which are written in Fortran
90 or Fortran 77. In total we analysed \SI{\overallLoC} lines of code
from 7 packages, of which we successfully parsed
\SI{\overalllinesParsed} lines. The ``Number of files'' column shows
how many files in each corpus that we were able to analyse with
Camfort. The most common reasons for Camfort rejecting a file were
either use of a C-preprocessor, or illegal use of language features
from a more modern Fortran variant.

\textbf{The Unified
  Model}~\cite{um}
is a weather forecasting and climate modelling tool developed by the
Met Office in the United Kingdom. It is used by research organisations
and meteorological services around the world. We use the development
branch (trunk) of the model. The code based in closed source but
institutional licenses are available for research purposes. The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model. We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.

% e3mg-ea
\textbf{E3MG} (An Energy-Environment-Economy (E3) Model at the Global Level) is a macroeconomic model used for assessment of environmental policy~\cite{RePEc:aen:journl:2006se-a12}. This model was developed by Cambridge Econometrics, an independent consultancy company.

\textbf{BLAS}~\cite{blas} (Basic Linear Algebra Subprograms) provides efficient and portable routines for vector and matrix operations. These routines feature in many other libraries (including LAPACK). We used version 3.6.0.

@@

\textbf{GEOS-Chem}~\cite{geos-chem} is a three-dimensional model of tropospheric chemistry developed at Harvard and used by approximately 70 universities and research institutions world-wide. We examined version 10-01.

\textbf{Navier} is a small project, giving a discrete
approximation to the two-dimensional Navier-Stokes fluid equations,
based on the book of~\citet{griebel1997numerical}.

@@

We first examined how frequently stencil computations occur in our
corpus. We define a \emph{potential} statement as one whose left-hand
side is an array subscript on neighbourhood indices. These statements
are the inputs to Step 2 of the Inference process described
Section~\ref{subsec:inference}. Each potential statement for which we
eventually generate some stencil specifications is also counted as
an \emph{actual} statement.

\begin{table}
\sisetup{table-format=4,table-text-alignment=right}
\footnotesize
\centering
\begin{tabular}{lS@{~~~}rS@{~~~}rSS[table-format=1]}
\hline
  & \multicolumn{2}{c}{Potential} & \multicolumn{2}{c}{Actual} & {Exact} & {Inexact}    \\
{Package}     & {\#} & \si{\percent} & {\#} & \si{\percent} \\
\hline
UM            & \umtickAssign & \SI{\umtickAssignPercent}{\percent} &
                \umtickAssignSuccess & \SI{\umtickAssignSuccessPercent}{\percent} &
                \umnumStencilLines &
                \umboundedBoth  \\
E3MG          & \ethreemgeatickAssign & \SI{\ethreemgeatickAssignPercent}{\percent} &
                \ethreemgeatickAssignSuccess & \SI{\ethreemgeatickAssignSuccessPercent}{\percent} &
                \ethreemgeanumStencilLines &
                \ethreemgeaboundedBoth \\
BLAS          & \blastickAssign & \SI{\blastickAssignPercent}{\percent} &
                \blastickAssignSuccess & \SI{\blastickAssignSuccessPercent}{\percent} &
                \blasnumStencilLines &
                \blasboundedBoth \\
Hybrid4       & \hybridfourtickAssign & \SI{\hybridfourtickAssignPercent}{\percent} &
                \hybridfourtickAssignSuccess & \SI{\hybridfourtickAssignSuccessPercent}{\percent} &
                \hybridfournumStencilLines &
                \hybridfourboundedBoth\\
GEOS-Chem     & \geoschemtickAssign & \SI{\geoschemtickAssignPercent}{\percent} &
                \geoschemtickAssignSuccess & \SI{\geoschemtickAssignSuccessPercent}{\percent} &
                \geoschemnumStencilLines &
                \geoschemboundedBoth \\
Navier        & \naviertickAssign & \SI{\naviertickAssignPercent}{\percent} &
                \naviertickAssignSuccess & \SI{\naviertickAssignPercent}{\percent} &
                \naviernumStencilLines &
                \navierboundedBoth \\
CP            & \computationalphysicstwotickAssign & \SI{\computationalphysicstwotickAssignPercent}{\percent} &
                \computationalphysicstwotickAssignSuccess & \SI{\computationalphysicstwotickAssignSuccessPercent}{\percent} &
                \computationalphysicstwonumStencilLines &
                \computationalphysicstwoboundedBoth \\
\hline
\textbf{Total}& \overalltickAssign & \SI{\overalltickAssignPercent}{\percent}  &
                \overalltickAssignSuccess & \SI{\overalltickAssignSuccessPercent}{\percent} &
                \overallnumStencilLines &
                \overallboundedBoth    \\
\hline
\end{tabular}
\caption{Stencil assignment rates\label{tab:tickAssign}}
\end{table}

The first two columns of Table~\ref{tab:tickAssign} shows the number
of potential and actual stencils and their rate as a percentage of the
total number of successfully parsed lines of code. There are a variety
of reasons why we might not infer an actual stencil from a potential
one: 1) \textbf{Non-subset induction variables} occurs when the
induction variables on the RHS are not a subset of the LHS. These
cases are not stencils by our definition. The degenereate case of this
is to have only constant indices on the LHS. We see lots of examples
of this in loops as accumulators e.g. computing the sum over some
other array; 2) \textbf{Derived induction variables} where the
indexing variable (\mintinline{fortran}{x}) is derived from an
induction variable (\mintinline{fortran}{i}) as in \mintinline{fortran}{x
  = len - i}; 3) \textbf{Inconsistent induction dimensions} occur when
an induction variable is used to specify more than one array dimension
on the RHS. These are common in matrix operations such as
LU-decomposition which contains assignments such as
\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.

@@

If the statement meets the stencil critera but the inference process cannot determine an exact stencil for a statement it will instead produce an upper (\term{atMost}) and lower (\term{atLeast}) bound. The number of exact and inexact stencils are shown in columns three and four of Figure \ref{tab:tickAssign}. We can see that our stencil language is capable of precisely specifying the overwhelming majority of stencils. Most of the inexact stencils apply to some form of data copying. For example (taken from E3MG):
\begin{minted}{fortran}
!- stencil atLeast, readOnce, irreflexive(dims=2), (reflexive(dim=1)) :: sfda
!- stencil atMost, readOnce, (forward(depth=8, dim=2))*(reflexive(dim=1)) :: sfdt
ZZ1(I,J) = SFDT(I,J+8)
\end{minted}

@@

We grouped stencils together into four classes based on similar levels of complexity. These four classes account for 99.4\% of all the stencils we found in the corpus.

@@

One relatively basic example of a stencil program is two dimensional
Jacobi iteration that repeatedly goes through each cell in a matrix
and computes the average value of the four adjacent cells. The kernel
can be expressed in a line of Fortran code,
\begin{minted}{fortran}
  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
\end{minted}
and Camfort is able to infer a precise stencil specification for this
line,
\begin{minted}{fortran}
  != stencil reflexive(dim=1)*centered(depth=1, dim=2, irreflexive) + reflexive(dim=2)*centered(depth=1, dim=1, irreflexive) :: a
\end{minted}
which when inserted back into the program can be checked by Camfort as
well. We then tested the error-detecting ability of Camfort by
generating a family of 6,561 Jacobi iteration programs by perturbing
the indices used in the following manner: each access of the array was
templated as \mintinline{fortran}{a(i+ki,j+kj)} and then all possible
combinations of \mintinline{fortran}{ki} and \mintinline{fortran}{kj}
were drawn from the set $\{-1,0,1\}$. There are four uses of
\mintinline{fortran}{ki} and four of \mintinline{fortran}{kj}, hence
$3^8=6561$ possible perturbations of this kind. This particular setup
tests the class of errors in which the programmer typed an incorrect
number but at least got the ordering of the indices correct, although
Camfort would also detect a mixup of variables in this case as well.

Camfort checked all 6,561 programs and found that only 24 of them were
valid implementations of the stencil specification. This was the
expected result because there are $4!$ possible correct permutations
of the four terms in the kernel. All of the other programs were found
to violate the specification in some way. Another way to test the
programs is through visual examination of their output in graphical
format. Figure~\ref{fig:jacobi} shows some of the results of running
the Jacobi iteration programs, rendered into image files. All of the
correct variants emitted images indistinguishable from
Figure~\ref{fig:jacobicorrect}, while all of the incorrect variants
had wildly different image outputs. Both the ``visual test'' and a
byte-by-byte comparison showed results that completely concurred with
the results of the Camfort test.

@@

@@

In this paper we have described an annotation language which can be
used to specify the access pattern of a stencil computation. We
defined a denotational semantic model for our language and proved its
soundness. This gave a basis for an equational theory of
specifications and also justifies our inference and checking
procedure. We evaluated our approach by implementing the stencil inference and
checking procedure in Camfort, an open-source analysis tool\footnote{\url{https://github.com/camfort/camfort}} for
Fortran programs and running the inference process over a corpus of
1.1 million lines of Fortran code.

@@

%%  LocalWords:  refactoring affine parameterised irreflexive atMost
%%  LocalWords:  centered readOnce atLeast discretisation Equational
%%  LocalWords:  equational disjunction denotational dimensionality
%%  LocalWords:  interprocedural Fortran Camfort preprocessor
%%  LocalWords:  committers

>>>>

\begin{abstract}
  Verifying the correctness of numerical computations relies on first
  giving a specification of their behaviour. In many cases, a
  numerical computation is derived from an underlying mathematical
  model, but this is often difficult to use as a specification as it
  is highly decoupled from the implementation. We have been studying
  specification techniques in the middle ground between low-level and
  high-level specifications for numerical code. In this paper, we
  focus on \emph{stencil computations}---a common idiom in numerical
  code, but one which is error-prone due to fine-grained indexing
  errors. We hypothesise that in practise, stencil computations tend
  to have a regular shape which is amenable to static analysis and
  simple specification. We describe a abstract spatial specification
  language for stencil computations that can be embedded as
  annotations in source code, along with inference, checking, and
  specification synthesis procedures. We evaluate our
  language against a corpus of numerical Fortran code, and show that
  the vast majority of stencil computations indeed have a simple,
  regular, static shape. Around 10\% of the stencils found evoke a
  programming pattern for which indexing errors are plausible. We
  investigate two stencils in detail and find that we are able to
  detect the vast majority of potential errors. This paper details our
  stencil specification language, its static semantics, inference,
  checking, and synthesis procedures, our implementation, and
  our evaluation study.
\end{abstract}

@@

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
(b) Inferred and synthesised specification for (a) by CamFort.
\vspace{0.5em}
\hrule
%
\begin{center}
\begin{tikzpicture}[scale=0.85]
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
\vspace{-0.5em}
(c) Pictorial representation of the two stencil specifications.
The horizontal dimension is \texttt{dim=1} and the vertical is \texttt{dim=2}.
\caption{Fragment of a Navier-Stokes fluid simulator and its
  specification in our language.}
\label{ref:navier-stokes-fragment}
\vspace{-0.8em}
\end{figure}

@@

\renewcommand*{\arraystretch}{0.8}
For the target language
syntax, $e$ ranges over its expressions (may be impure) and $v$
overs its (imperative) variables. %By
%``variable'', we mean in the imperative sense (named binders to memory cells).

\begin{definition}[Base induction variable]
  A variable \textit{v} of integer type is a \emph{base induction
    variable} if it is the control variable of a ``for''
  loop,\footnote{Or equivalent, \eg{}, \texttt{do} in Fortran, for our implementation.} incremented by $1$ per
iteration. The variable is marked as an induction variable
only within the body of the loop. Since we ignore non-base
(derived) induction variables, we simply say \emph{induction
  variable}. Induction variables are ranged over by $i$, $j$, $k$ throughout. 
\end{definition}

@@

\begin{figure}[t]
\vspace{-0.9em}
\input{region-model.tex}
%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
%  the multi-set $A$}
\caption{Model of regions,
$\interp{-}_n : \textit{region} \rightarrow \mathcal{P}(\domainVal{}^n)$}
\label{fig:region-model}
\vspace{-0.7em}
\end{figure}

@@

\begin{definition} \textsf{Mult}
and \textsf{Approx} are parametric labelled variant types
with injection function given by their definition:
%
\begin{align*}
\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
\end{align*}
\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
on. The $\textsf{Mult}$ type corresponds to the presence or absence of the
\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}. 
The $\textsf{Approx}$ type corresponds to the presence or absence of
the spatial approximation modifier, with \textsf{exact} when there is
no such modifier and \textsf{lower} and \textsf{upper} for
\term{atLeast} and \term{atMost}. Later, it will be useful to model
pairs of lower and upper bounds which is provided by \textsf{both} 
(but which isn't used here). These are used for labelled models, and other components later in
Section~\ref{sec:analysis}.% with additional information.
\label{def:mult-and-approx}
\end{definition}
\vspace{-0.5em}

\begin{figure}
\vspace{-0.5em}
\input{spatial-model.tex}
\caption{\textit{spec} model,
$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
\label{fig:spatial-model}
\vspace{-0.75em}
\end{figure}

\subsection{Soundness}
%
Given the above model, we can
consider soundness for the equational theory of
specifications and the notion of spatial approximation. The proofs are given in our supplement.

@@

\begin{definition}[Approximation]
A specification $S'$
\emph{approximates} a more specific specification $S$, written 
$S \sqsubseteq S'$, \emph{iff}:
$$
\forall M_c \in \textsf{Mult}(\domainVal{}^n) . \mathit{cons}_n(\interp{S}_n,M_c) \implies
\mathit{cons}(\interp{S'}_n,M_c)
$$
\end{definition}
%
For the syntactic sub-specification equations on regions $<:^r$, it follows that
these imply semantic approximation:
%
\begin{lemma}[Sub-specification $<:^r$ coincides with approx 
$\sqsubseteq$]
\begin{equation*}
\forall S, S' . \; S <:^r S' \Rightarrow S \sqsubseteq S'
\end{equation*}
\end{lemma}
%
\noindent
However, the converse direction does not hold. 
The regions describing the more acceptiong five-point stencil
above do not follow from the definition of $<:$ on the previous
more specific specification.%, that is:
%${\textcolor{darkgreen}{\term{centered(depth=1, dim=1)*reflexive(dim=2)
%      + centered(depth=1, dim=2)*reflexive(dim=1)}}}
%\not{<:} {\textcolor{darkgreen}{\term{centered(depth=1,dim=1) +
%      centered(depth=1,dim=2)}}}$.
 However, consitency (and thus $\sqsubseteq$) shows
that we can check the five-point stencil code against the more general
specifcation of {\textcolor{darkgreen}{\term{fivepoint}}}. Thus, we
can always use an approximation in place of the
more specific specification. Although this decreases the likelihood of catching
stencil errors (as more access patterns conform), it reduces the space and
effort it takes to write the specification.

\section{Evaluation}
\label{sec:evaluation}
We evaluated our specificataion language on a corpus of scientific
code.  We overview the corpus itself before presenting our results.

@@

Package   & {Total} & {Parsed} & {Total} & {Parsed}  \\ % braces prevent headers from being formatted as numbers
\hline
UM        & \umFiles                       & \umparseOk                      & \umLoC                      & \umlinesParsed \\
E3MG      & \ethreemgeaFiles               & \ethreemgeaparseOk              & \ethreemgeaLoC              & \ethreemgealinesParsed \\
BLAS      & \blasFiles                     & \blasparseOk                    & \blasLoC                    & \blaslinesParsed \\
Hybrid4   & \hybridfourFiles               & \hybridfourparseOk              & \hybridfourLoC              & \hybridfourlinesParsed \\
GEOS-Chem & \geoschemFiles                 & \geoschemparseOk                & \geoschemLoC                & \geoschemlinesParsed \\
Navier    & \navierFiles                   & \navierparseOk                  & \navierLoC                  & \navierlinesParsed \\
CP        & \computationalphysicstwoFiles  & \computationalphysicstwoparseOk & \computationalphysicstwoLoC & \computationalphysicstwolinesParsed \\
\hline
\textbf{Total} & \overallFiles  & \overallparseOk & \overallLoC & \overalllinesParsed \\
\hline
\end{tabular}
\caption{Summary of software packages used for
  evaluation\label{tab:corpus}}
\vspace{-1em}
\end{table}

Table~\ref{tab:corpus} shows summary statistics of the software
packages used in our evaluation, all of which are written in Fortran
90 or Fortran 77. In total we analysed \SI{\overallLoC} lines of code
from 7 packages, of which we successfully parsed
\SI{\overalllinesParsed} lines. The ``Number of files'' column shows
how many files in each corpus that we were able to analyse with
CamFort. The most common reasons for CamFort rejecting a file were
either use of a C-preprocessor, or illegal use of language features
from a modern Fortran variant.

\textbf{The Unified
  Model}~\cite{um}
is a weather forecasting and climate modelling tool developed by the
Met Office in the United Kingdom. It is used by research organisations
and meteorological services around the world. We use the development
branch (trunk) of the model. The code base in closed source but
institutional licenses are available for research purposes. The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model. We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.

% e3mg-ea
\textbf{E3MG} (An Energy-Environment-Economy (E3) Model at the Global Level) is a macroeconomic model used for assessment of environmental policy~\cite{RePEc:aen:journl:2006se-a12}. This was developed by Cambridge Econometrics, an independent consultancy company.

\textbf{BLAS}~\cite{blas} (Basic Linear Algebra Subprograms) is a popular
library providing efficient and portable routines for vector and
matrix operations. These routines feature in many other libraries
(including LAPACK). We used version 3.6.0. We chose to include this
package for breadth, as it provides general numerical functions rather
than a specialised scientific model.

@@

\textbf{GEOS-Chem}~\cite{geos-chem} is a three-dimensional model of tropospheric chemistry developed at Harvard and used by $\sim$70 universities and research institutions world-wide. We examined version 10-01.

\textbf{Navier} is a small numerical simulation, giving a discrete
approximation to the two-dimensional Navier-Stokes fluid equations,
based on the book of~\citet{griebel1997numerical}.

@@

We first examined how frequently stencil computations occur in our
corpus. This was carried out by running our implementation in 
whole-code-base inference mode. We define a \emph{potential} statement as one whose left-hand
side is an array subscript on neighbourhood indices. These statements
are the inputs to Step 2 of the Inference process described in 
Section~\ref{subsec:inference}. Each potential statement for which we
eventually generate some specification is also counted as
an \emph{actual} statement.

\begin{table}
\sisetup{table-format=4,table-text-alignment=right}
\footnotesize
\centering
\begin{tabular}{lS@{~~~}rS@{~~~}rSS[table-format=1]}
\hline
  & \multicolumn{2}{c}{Potential} & \multicolumn{2}{c}{Actual} & {Exact} & {Inexact}    \\
{Package}     & {\#} & \si{\percent} & {\#} & \si{\percent} \\
\hline
UM            & \umtickAssign & \SI{\umtickAssignPercent}{\percent} &
                \umtickAssignSuccess & \SI{\umtickAssignSuccessPercent}{\percent} &
                \umnumStencilLines &
                \umboundedBoth  \\
E3MG          & \ethreemgeatickAssign & \SI{\ethreemgeatickAssignPercent}{\percent} &
                \ethreemgeatickAssignSuccess & \SI{\ethreemgeatickAssignSuccessPercent}{\percent} &
                \ethreemgeanumStencilLines &
                \ethreemgeaboundedBoth \\
BLAS          & \blastickAssign & \SI{\blastickAssignPercent}{\percent} &
                \blastickAssignSuccess & \SI{\blastickAssignSuccessPercent}{\percent} &
                \blasnumStencilLines &
                \blasboundedBoth \\
Hybrid4       & \hybridfourtickAssign & \SI{\hybridfourtickAssignPercent}{\percent} &
                \hybridfourtickAssignSuccess & \SI{\hybridfourtickAssignSuccessPercent}{\percent} &
                \hybridfournumStencilLines &
                \hybridfourboundedBoth\\
GEOS-Chem     & \geoschemtickAssign & \SI{\geoschemtickAssignPercent}{\percent} &
                \geoschemtickAssignSuccess & \SI{\geoschemtickAssignSuccessPercent}{\percent} &
                \geoschemnumStencilLines &
                \geoschemboundedBoth \\
Navier        & \naviertickAssign & \SI{\naviertickAssignPercent}{\percent} &
                \naviertickAssignSuccess & \SI{\naviertickAssignPercent}{\percent} &
                \naviernumStencilLines &
                \navierboundedBoth \\
CP            & \computationalphysicstwotickAssign & \SI{\computationalphysicstwotickAssignPercent}{\percent} &
                \computationalphysicstwotickAssignSuccess & \SI{\computationalphysicstwotickAssignSuccessPercent}{\percent} &
                \computationalphysicstwonumStencilLines &
                \computationalphysicstwoboundedBoth \\
\hline
\textbf{Total}& \overalltickAssign & \SI{\overalltickAssignPercent}{\percent}  &
                \overalltickAssignSuccess & \SI{\overalltickAssignSuccessPercent}{\percent} &
                \overallnumStencilLines &
                \overallboundedBoth    \\
\hline
\end{tabular}
\caption{Stencil assignment rates\label{tab:tickAssign}}
\vspace{-0.8em}
\end{table}

The first two columns of Table~\ref{tab:tickAssign} show the number
of potential and actual stencils and their rate as a percentage of the
total number of successfully parsed lines of code. The ratio of
actual to potential stencils overall is thus $32\%$, \ie{}, we give
a specification to roughly a third of the potential stencil specifications.

There are a variety of reasons why we might not infer a actual specification from a potential
stencil computation: \\
\noindent 
1) \textbf{Non-subset induction variables} occur when the
induction variables on the RHS are not a subset of those in the LHS. These
cases are not stencils by our definition. The degenereate case of this
is to have only constant indices on the LHS. We see lots of examples
of this in loops as accumulators \eg{} computing the sum over an array; \\
\noindent
 2) \textbf{Derived induction variables} where the
indexing variable (\mintinline{fortran}{x}) is derived from an
induction variable (\mintinline{fortran}{i}) as in
\mintinline{fortran}{x = len - i};  \\
\noindent
3) \textbf{Inconsistent induction dimensions} occur when
an induction variable is used to specify more than one array dimension
on the RHS or multiple induction variables are used for the same
dimension on the RHS. These are common in matrix operations such as
LU-decomposition with assignments such as
\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.

@@

If the statement meets the stencil critera but the inference process
cannot determine an exact stencil for a statement it will instead
produce an upper (\term{atMost}) and lower (\term{atLeast}) bound. The
number of exact and inexact stencils are shown in columns three and
four of Table~\ref{tab:tickAssign}. Note that the total for ``Exact``
is higher than ``Actual`` because for some lines we produce multiple
specifications (perhaps for different array variables).
We see that our specification language is capable of precisely specifying the overwhelming majority of stencils. Most of the inexact stencils apply to some form of data copying. For example (taken from E3MG):
\begin{minted}{fortran}
!- stencil atLeast, readOnce, irreflexive(dims=2), (reflexive(dim=1)) :: sfda
!- stencil atMost, readOnce, (forward(depth=8, dim=2))*(reflexive(dim=1)) :: sfdt
ZZ1(I,J) = SFDT(I,J+8)
\end{minted}

@@

We grouped our inferred specifications together into four classes based on similar levels of complexity. These four classes account for 99.4\% of all the stencils we found in the corpus.

@@

One relatively basic example of a stencil program is two dimensional
Jacobi iteration that repeatedly goes through each cell in a matrix
and computes the average value of the four adjacent cells. The kernel
can be expressed in a line of Fortran code,
\begin{minted}{fortran}
  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
\end{minted}
and CamFort is able to infer a precise stencil specification for this
line,
\begin{minted}{fortran}
  != stencil reflexive(dim=1)*centered(depth=1, dim=2, irreflexive) + reflexive(dim=2)*centered(depth=1, dim=1, irreflexive) :: a
\end{minted}
which when inserted back into the program can be checked by CamFort as
well. We then tested the error-detecting ability of CamFort by
generating a family of 6,561 Jacobi iteration programs by perturbing
the indices used in the following manner: each access of the array was
templated as \mintinline{fortran}{a(i+ki,j+kj)} and then all possible
combinations of \mintinline{fortran}{ki} and \mintinline{fortran}{kj}
were drawn from the set $\{-1,0,1\}$. There are four uses of
\mintinline{fortran}{ki} and four of \mintinline{fortran}{kj}, hence
$3^8=6561$ possible perturbations of this kind. This particular setup
tests the class of errors in which the programmer typed an incorrect
number but at least got the ordering of the indices correct, although
CamFort would also detect a mixup of variables in this case as well.

CamFort checked all 6,561 programs and found that only 24 of them were
valid implementations of the stencil specification. This was the
expected result because there are $4!$ possible correct permutations
of the four terms in the kernel. All of the other programs were found
to violate the specification in some way. Another way to test the
programs is through visual examination of their output in graphical
format. Figure~\ref{fig:jacobi} shows some of the results of running
the Jacobi iteration programs, rendered into image files. All of the
correct variants emitted images indistinguishable from
Figure~\ref{fig:jacobicorrect}, while all of the incorrect variants
had wildly different image outputs. Both the ``visual test'' and a
byte-by-byte comparison showed results that completely concurred with
the results of the CamFort test.

@@

\paragraph{Program synthesis}

@@

In this paper we have described an annotation language which can be
used to specify the access pattern of a stencil computation. We
defined a denotational semantic model for our language and proved its
soundness. This gave a basis for an equational theory of
specifications and also justifies our inference and checking
procedure. We evaluated our approach by implementing the stencil inference and
checking procedure in CamFort, an open-source analysis tool\footnote{\url{https://github.com/camfort/camfort}} for
Fortran programs and running the inference process over a corpus of
1.1 million lines of Fortran code.

@@

%%  LocalWords:  refactoring affine parameterised irreflexive atMost
%%  LocalWords:  centered readOnce atLeast discretisation Equational
%%  LocalWords:  equational disjunction denotational dimensionality
%%  LocalWords:  interprocedural Fortran CamFort preprocessor
%%  LocalWords:  committers
